---
title: "translocation analysis - orientation"
author: "steph"
date: "2024-09-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

load packages, read in processed dataset, normalize variables

```{r}
library(tidyverse)
library(NatParksPalettes)
library(viridis)
library(ggpubr)
library(readxl)
library(lme4)
library(car)

theme_set(theme_classic())

#edit to quantNorm function - remove NAs from estimation
quantNorm =function(x){
  x1<-x[!is.na(x)]
  x[!is.na(x)]<-qnorm(rank(x1,ties.method = "average")/(length(x1)+1))
  x
}


```

```{r}


birds<-read_xlsx("C:/Users/Steph/TAMU_OneDrive/Thrushes/translocation/data_archiving/translocation_data.xlsx",sheet=2)
pembertonBirds<-read_xlsx("C:/Users/Steph/TAMU_OneDrive/Thrushes/translocation/data_archiving/translocation_data.xlsx",sheet=3)


birds2<-rbind(pembertonBirds,birds)%>%
  mutate(fall_releaseSite_bearing_tier1=as.numeric(fall_releaseSite_bearing_tier1))%>%
  mutate(norm_releaseSite_bearing_1=quantNorm(fall_releaseSite_bearing_tier1))%>%
  mutate(ancestry=aims_ancestry*(-1)+1)

birds<-birds%>%
  mutate(fall_releaseSite_bearing_tier1=as.numeric(fall_releaseSite_bearing_tier1))%>%
  mutate(norm_releaseSite_bearing_1=quantNorm(fall_releaseSite_bearing_tier1))%>%
  mutate(ancestry=aims_ancestry*(-1)+1)


```

### Orientations from Lilloet and Squamish

* expect effects of release site (environment), ancestry (genetics), and interaction between release site and environment (genetics x environment)
* orientation estimated in two ways: relative to the hybrid zone (i.e. Pemberton) and relative to each bird's release location

```{r,fig.width=5,fig.height=3}

ggplot(birds,aes(x=ancestry,y=norm_releaseSite_bearing_1,
                 group=release_site,
                 colour=release_site))+
  geom_smooth(method="lm",alpha=0.18)+
  geom_point(size=4,shape=19,alpha=0.5)+
  scale_color_manual(values=natparks.pals("Banff")[c(4,1)],name="Release site")+
  xlab("Ancestry")+
  ylab("Orientation to\nrelease site")

```


```{r,fig.width=4,fig.height=3}

lm2<-lm(norm_releaseSite_bearing_1~release_site*ancestry,data=birds,
        na.action=na.exclude)
anova(lm2)
summary(lm2)
emmeans::emmeans(lm2, ~release_site)

p=visreg::visreg(lm2,"release_site",by="ancestry",overlay=T,breaks=c(0.25,0.5,0.75))

gg1<-ggplot(p$fit, aes(release_site, visregFit,
                  colour=factor(ancestry),group=as.factor(ancestry))) +
  geom_errorbar(aes(ymin=visregLwr, ymax=visregUpr), alpha=0.5, 
              linetype=1, size=0.2,width=0.1,position=position_dodge(width=0.2)) +
  geom_point(position=position_dodge(width=0.2),size=2) +
  geom_line(position=position_dodge(width=0.2),size=0.8)+
  scale_color_manual(values=natparks.pals("Banff")[c(1,6,4)],name="migratory\ngenotype")+
  labs(fill="Weight")+
  ylab("orientation")+xlab("release site")
gg1

```

### Comparison to Pemberton birds

* selected birds released in Pemberton in 2020 & 2023
* bearing shows a negative relationship to ancestry - inland birds fly east, coastal birds fly west


```{r,fig.width=9,fig.height=3}

p1<-ggplot(birds,aes(x=ancestry,y=norm_releaseSite_bearing_1,
                 group=release_site,
                 colour=release_site))+
  geom_smooth(method="lm",alpha=0.18)+
  geom_point(size=4,shape=19,alpha=0.5)+
  scale_color_manual(values=natparks.pals("Banff")[c(4,1)],name="Release site")+
  xlab("Ancestry")+
  ylab("Orientation")

p2<-ggplot(birds2,aes(x=ancestry,y=norm_releaseSite_bearing_1,
                 group=release_site,
                 colour=release_site))+
  geom_smooth(method="lm",alpha=0.18)+
  geom_point(size=4,shape=19,alpha=0.5)+
  scale_color_manual(values=natparks.pals("Banff")[c(4,7,1)],name="Release site")+
  xlab("Ancestry")+
  ylab("Orientation")

ggarrange(p1,p2,labels=LETTERS)

#ggsave("C:/Users/Steph/TAMU_OneDrive//Thrushes/translocation/FigS1.png",
#       width=9,height=3)

```


* include year as a random effect, to control for similarities among 2020 vs 2023 birds
* here, release site is unimportant, but ancestry is significant and there is an interaction between release site and ancestry
* again, more positive slopes between release site and ancestry in Lilloet and Squamish than Pemberton

```{r,fig.width=5,fig.height=4}

birds2<-birds2%>%mutate(f_release_year=as.factor(release_year))%>%
  mutate(f_release_site=factor(release_site,levels=c("Squamish","Pemberton","Lilloet")))

lmer2<-lmer(norm_releaseSite_bearing_1~f_release_site*ancestry+(1|f_release_year),
            data=birds2,na.action=na.exclude)
Anova(lmer2)

coef(lmer2)
summary(lmer2)

data.frame(Anova(lmer2))%>%
  mutate()
```

```{r,fig.width=5.5,fig.height=4}


p=visreg::visreg(lmer2,"f_release_site",by="ancestry",overlay=T,breaks=c(0.25,0.5,0.75))

gg1<-ggplot(p$fit, aes(f_release_site,
                  visregFit,
                  colour=factor(ancestry),group=as.factor(ancestry))) +
  geom_errorbar(aes(ymin=visregLwr, ymax=visregUpr), 
              linetype=1, size=0.6,width=0.1,position=position_dodge(width=0.2)) +
  geom_point(position=position_dodge(width=0.2),size=3) +
  geom_line(position=position_dodge(width=0.2),size=1)+
  scale_color_manual(values=natparks.pals("Banff")[c(1,7,5)],name="Ancestry",
                     labels=c("Coastal backcross","Hybrid","Inland backcross"))+
  labs(fill="Weight")+
  ylab("Orientation")+xlab("Release site")

gg1
#ggsave(plot=gg1,
#       file="C:/Users/Steph/TAMU_OneDrive/Thrushes/translocation/Fig3B.png",
#       width=5.5,height=4)

```

