---
title: "translocation analysis survival"
author: ""
date: "2025-09-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

* load packages & read in processed dataset
* adjust ancestry so that *coastal = 0* and *inland = 1*

```{r}
library(tidyverse)
library(NatParksPalettes)
library(viridis)
library(ggpubr)
library(readxl)
library(scico)

theme_set(theme_classic())

#manual change made to CJS output - deleted header lines to allow output to be read into R
cjs_out<-read.table("C:/Users/Steph/GitHub_data/translocation/CJS.beta.translocation.20250914.edit.txt",
           header=T)
cjs_out_pemberton<-read.table("C:/Users/Steph/GitHub_data/translocation/CJS.beta.translocation_Pemberton.20250914.edit.txt",
           header=T)

phi<-read.csv("C:/Users/Steph/GitHub_data/translocation/CJS.fit.phi.translocation.20250914.csv")
phi_pemberton<-read.csv("C:/Users/Steph/GitHub_data/translocation/CJS.fit.phi.translocation_Pemberton.20250914.csv")

birds<-read_xlsx("C:/Users/Steph/TAMU_OneDrive/Thrushes/translocation/data_archiving/translocation_data_revision.xlsx",sheet=2)
pembertonBirds<-read_xlsx("C:/Users/Steph/TAMU_OneDrive/Thrushes/translocation/data_archiving/translocation_data_revision.xlsx",sheet=3)

birds2<-rbind(pembertonBirds,birds)%>%
  mutate(ancestry=hiest_ancestry*(-1)+1)

birds<-birds%>%
  mutate(ancestry=hiest_ancestry*(-1)+1)

```

```{r}

tabS1<-birds2%>%
  mutate(ancestry=round(ancestry,3),
    capture_gps.n=round(capture_gps.n,3),
    capture_gps.w=round(capture_gps.w,3),
    release_gps.n=round(release_gps.n,3),
    release_gps.w=round(release_gps.w,3),
    release_site=if_else(release_site=="Lilloet","Lillooet",release_site))%>%
  dplyr::select(reference,translocated,ancestry,release_site,
    capture_gps.n,capture_gps.w,release_gps.n,
    release_gps.w,release_date)
#tabS1
#write_csv(tabS1,
#         "C:/Users/Steph/TAMU_OneDrive/Thrushes/translocation/fig_drafts/TableS1.csv")

rm(tabS1)

```


```{r}



poly.all<-data.frame(X.co=rep(c(0,0.5,1)), #make triangle polygon for plotting
                     Y.co=rep(c(0,1,0)))

figTri<-ggplot(birds2%>%mutate(release_site=if_else(release_site=="Lilloet","Lillooet",release_site)),
       aes(x=ancestry,y=hiest_heterozygosity,colour=release_site))+
  geom_polygon(data=poly.all,aes(x=X.co,y=Y.co),fill=NA,colour="grey30")+
  geom_point(size=4,shape=19,alpha=0.7)+
  scale_color_manual(values=scico(n=3,palette='navia',end=0.8),
                     guide="none")+
  coord_equal()+
  xlim(0,1)+ylim(0,1)+
  facet_wrap(vars(release_site))+
  xlab("Ancestry - coastal to inland")+
  ylab("Interspecific heterozygosity")
figTri
ggsave(plot=figTri,
       "C:/Users/Steph/TAMU_OneDrive/Thrushes/translocation/FigS_tri.png",
       width=6,height=3)

```

* Sample sizes for survival analysis

```{r}

knitr::kable(birds2%>%
               group_by(release_site,release_year)%>%
               summarise(count=n()))

```


### CJS (mark recapture) models

* Fit a CJS model with detections binned into 10 day chunks to create 30 time points (data were too sparse with each day as a time point)
* Model: Survival ~ release_site + ancestry + release_site x ancestry; Detection ~ time + release_site + ancestry
* No evidence for the predicted interaction between release site and ancestry
* None of release site, ancestry, or the interaction between release site and ancestry had a significant effect on survival
* probability of detection varied among time points, with the higher probabilities of detection occurring during fall and spring migration
* birds released in Squamish had a higher probability of detection than those released in Lilloet, but detection probability did not vary with ancestry
* Once you control for difference in detection probability - no difference in survival between release sites

Survival analysis results:

```{r}

knitr::kable(cjs_out%>%filter(startsWith(Factor,"Phi"))%>%
  mutate_if(is.numeric,round,digits=2)%>%
    select(-se)%>%
    rename(Lower_CL=lcl,Upper_CL=ucl))

```

Detection probability analysis results:

(minus variation among time points)

```{r}

knitr::kable(cjs_out%>%filter(startsWith(Factor,"p."))%>%
               filter(!startsWith(Factor,"p.time"))%>%
  mutate_if(is.numeric,round,digits=2)%>%
    select(-se)%>%
    rename(Lower_CL=lcl,Upper_CL=ucl))

```

### CJS - with Pemberton

* Here, I would expect a flat relationship between ancestry and survival in Pemberton (ie at the centre of the hybrid zone), a positive relationship in Lilloet (better survival of inland birds) and a negative relationship in Squamish (better survival of coastal birds)
* Model: Survival ~ release_site + ancestry + release_site x ancestry; Detection ~ time + release_site + release_year + ancestry
* Adding Pemberton birds tagged in 2020 & 2023 does not change anything for the survival results - everything is still null
* Because release site and release year aren't entirely covarying, I added relase year (alongside time point, release site, and ancestry) as a predictor for detection probability

Survival analysis results:

```{r}

knitr::kable(cjs_out_pemberton%>%filter(startsWith(Factor,"Phi"))%>%
  mutate_if(is.numeric,round,digits=2)%>%
    select(-se)%>%
    rename(Lower_CL=lcl,Upper_CL=ucl))

```

Detection probability analysis results:

```{r}

knitr::kable(cjs_out_pemberton%>%filter(startsWith(Factor,"p."))%>%
               filter(!startsWith(Factor,"p.time"))%>%
  mutate_if(is.numeric,round,digits=2)%>%
    select(-se)%>%
    rename(Lower_CL=lcl,Upper_CL=ucl))

```

```{r}

library(lme4)
library(car)

glm1<-glm(survival~ancestry*release_site,
      family = binomial(link="logit"),
      data=birds)

summary(glm1)
Anova(glm1)

glmm1<-glmer(survival~ancestry*release_site+(1|release_year),
      family = binomial(link="logit"),
      data=birds2)

summary(glmm1)
Anova(glmm1)



```

### Survival figures

* slopes are based on ancestry x release site predictions from the CJS models
* I think I actually prefer the violin plots because the data are more visible

```{r,fig.width=6,fig.height=3}

ggplot(birds,aes(x=ancestry,y=survival,
                 group=release_site,
                 colour=release_site))+
  #geom_smooth(method="lm",alpha=0.18)+
  geom_point(size=4,shape=19,alpha=0.8,position=position_jitter(width=0,height=0.1))+
  geom_line(data=phi,aes(x=ancestry,y=est.adj,colour=release_site),size=1)+
  scale_color_manual(values=scico(n=3,palette='navia',end=0.8)[c(3,1)])+
  xlab("ancestry")+
  ylab("survival")


```

```{r,fig.width=6,fig.height=3}


ggplot(birds2,aes(x=ancestry,y=survival,
                 group=release_site,
                 colour=release_site))+
  geom_point(size=4,shape=19,alpha=0.8,position=position_jitter(width=0,height=0.1))+
  geom_line(data=phi_pemberton,
            aes(x=ancestry,y=est.adj,colour=release_site),size=1)+
  scale_color_manual(values=natparks.pals("Banff")[c(4,6,1)])+
  xlab("ancestry")+
  ylab("survival")


```

```{r}

birds2$release_site<-factor(birds2$release_site,levels=c("Squamish","Pemberton","Lilloet"))

```

```{r,fig.width=6,fig.height=4}


gg1<-ggplot(birds2,aes(y=ancestry,x=release_site,
                 group=interaction(as.factor(survival),release_site),
                 colour=release_site,shape=as.factor(survival)))+
  geom_point(size=3,alpha=0.8,
             position=position_jitterdodge(
               jitter.width =0.2,jitter.height=0,dodge.width=0.9))+
  scale_shape_manual(values=c(1,19),name="",labels=c("Not survived","Survived"))+
  scale_color_manual(values=scico(n=3,palette='navia',end=0.8),
                     guide="none")+
  geom_violin(fill=NA,aes(group=interaction(survival,release_site)),size=0.8)+
  ylab("Ancestry")+xlab("Release site")+ylim(0,1)+
  scale_x_discrete(labels=c("Squamish","Pemberton","Lillooet"))
  
gg1

#ggsave("C:/Users/Steph/TAMU_OneDrive/Thrushes/translocation/Fig2.svg",
#       plot=gg1,width=6,height=4)

```

* Next, points coloured by model-fitted survival probabilities

```{r,fig.width=5,fig.height=4}


ggplot(phi,aes(x=release_site,y=ancestry,colour=est.adj))+
  geom_jitter(size=4,shape=19,alpha=0.8,width=0.1)+
  viridis::scale_colour_viridis(option="mako",end=0.9,name="survival\nprobability")


```

```{r,fig.width=6,fig.height=4}


ggplot(phi_pemberton,aes(x=release_site,y=ancestry,colour=est.adj))+
  geom_jitter(size=4,shape=19,alpha=0.8,width=0.2)+
  viridis::scale_colour_viridis(option="mako",end=0.9,name="survival\nprobability")


```