---
title: "translocation analysis survival"
author: "steph"
date: "2024-09-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

* load packages & read in processed dataset
* adjust ancestry so that *coastal = 0* and *inland = 1*

```{r}
library(tidyverse)
library(NatParksPalettes)
library(viridis)
library(ggpubr)
library(readxl)

theme_set(theme_classic())

#manual change made to CJS output - deleted header lines to allow output to be read into R
cjs_out<-read.table("C:/Users/Steph/GitHub_data/translocation/CJS.beta.translocation.20250410.edit.txt",
           header=T)
cjs_out_pemberton<-read.table("C:/Users/Steph/GitHub_data/translocation/CJS.beta.translocation_Pemberton.20250410.edit.txt",
           header=T)

phi<-read.csv("C:/Users/Steph/GitHub_data/translocation/CJS.fit.phi.translocation.20250410.csv")
phi_pemberton<-read.csv("C:/Users/Steph/GitHub_data/translocation/CJS.fit.phi.translocation_Pemberton.20250410.csv")

birds<-read_xlsx("C:/Users/Steph/TAMU_OneDrive/Thrushes/translocation/data_archiving/translocation_data.xlsx",sheet=2)
pembertonBirds<-read_xlsx("C:/Users/Steph/TAMU_OneDrive/Thrushes/translocation/data_archiving/translocation_data.xlsx",sheet=3)

birds2<-rbind(pembertonBirds,birds)%>%
  mutate(ancestry=aims_ancestry*(-1)+1)

birds<-birds%>%
  mutate(ancestry=aims_ancestry*(-1)+1)
birds2$capture_gps.n
write.csv(birds2%>%
            mutate(capture_gps.n=round(capture_gps.n,3),
                   capture_gps.w=round(capture_gps.w,3),
                   release_gps.n=round(release_gps.n,3),
                   release_gps.w=round(release_gps.w,3))%>%
            dplyr::select(reference,translocated,release_site,
                          capture_gps.n,capture_gps.w,release_gps.n,
                          release_gps.w,release_date),
          "C:/Users/Steph/TAMU_OneDrive/Thrushes/translocation/fig_drafts/TableS1.csv",
          row.names=F)

```

* Sample sizes for survival analysis

```{r}

knitr::kable(birds2%>%
               group_by(release_site,release_year)%>%
               summarise(count=n()))

```


### CJS (mark recapture) models

* Fit a CJS model with detections binned into 10 day chunks to create 30 time points (data were too sparse with each day as a time point)
* Model: Survival ~ release_site + ancestry + release_site x ancestry; Detection ~ time + release_site + ancestry
* No evidence for the predicted interaction between release site and ancestry
* None of release site, ancestry, or the interaction between release site and ancestry had a significant effect on survival
* The direction of the interaction effect was not in the predicted direction - the release site x ancestry interaction is positive, meaning that the relationship between ancestry and survival is more positive for birds released in Squamish relative to those released in Lilloet
* probability of detection varied among time points, with the higher probabilities of detection occurring during fall and spring migration
* birds released in Squamish had a higher probability of detection than those released in Lilloet, but detection probability did not vary with ancestry
* Once you control for difference in detection probability - no difference in survival between release sites

Survival analysis results:

```{r}

knitr::kable(cjs_out%>%filter(startsWith(Factor,"Phi"))%>%
  mutate_if(is.numeric,round,digits=2)%>%
    select(-se)%>%
    rename(Lower_CL=lcl,Upper_CL=ucl))

```

Detection probability analysis results:

(minus variation among time points)

```{r}

knitr::kable(cjs_out%>%filter(startsWith(Factor,"p."))%>%
               filter(!startsWith(Factor,"p.time"))%>%
  mutate_if(is.numeric,round,digits=2)%>%
    select(-se)%>%
    rename(Lower_CL=lcl,Upper_CL=ucl))

```

### CJS - with Pemberton

* Here, I would expect a flat relationship between ancestry and survival in Pemberton (ie at the centre of the hybrid zone), a positive relationship in Lilloet (better survival of inland birds) and a negative relationship in Squamish (better survival of coastal birds)
* Model: Survival ~ release_site + ancestry + release_site x ancestry; Detection ~ time + release_site + release_year + ancestry
* Adding Pemberton birds tagged in 2020 & 2023 does not change anything for the survival results - everything is still null
* Because release site and release year aren't entirely covarying, I added relase year (alongside time point, release site, and ancestry) as a predictor for detection probability
* For detection probabilities, the difference seems to be primarily between release years, but also a reduction in Lilloet relative to Pemberton - not sure how much to read into that

Survival analysis results:

```{r}

knitr::kable(cjs_out_pemberton%>%filter(startsWith(Factor,"Phi"))%>%
  mutate_if(is.numeric,round,digits=2)%>%
    select(-se)%>%
    rename(Lower_CL=lcl,Upper_CL=ucl))

```

Detection probability analysis results:

```{r}

knitr::kable(cjs_out_pemberton%>%filter(startsWith(Factor,"p."))%>%
               filter(!startsWith(Factor,"p.time"))%>%
  mutate_if(is.numeric,round,digits=2)%>%
    select(-se)%>%
    rename(Lower_CL=lcl,Upper_CL=ucl))

```

### Survival figures

* slopes are based on ancestry x release site predictions from the CJS models
* I think I actually prefer the violin plots because the data are more visible

```{r,fig.width=6,fig.height=3}

ggplot(birds,aes(x=ancestry,y=survival,
                 group=release_site,
                 colour=release_site))+
  #geom_smooth(method="lm",alpha=0.18)+
  geom_point(size=4,shape=19,alpha=0.8,position=position_jitter(width=0,height=0.1))+
  geom_line(data=phi,aes(x=ancestry,y=est.adj,colour=release_site),size=1)+
  scale_color_manual(values=natparks.pals("Banff")[c(4,1)])+
  xlab("ancestry")+
  ylab("survival")


```

```{r,fig.width=6,fig.height=3}


ggplot(birds2,aes(x=ancestry,y=survival,
                 group=release_site,
                 colour=release_site))+
  geom_point(size=4,shape=19,alpha=0.8,position=position_jitter(width=0,height=0.1))+
  geom_line(data=phi_pemberton,
            aes(x=ancestry,y=est.adj,colour=release_site),size=1)+
  scale_color_manual(values=natparks.pals("Banff")[c(4,6,1)])+
  xlab("ancestry")+
  ylab("survival")


```


```{r,fig.width=6,fig.height=4}

gg1<-ggplot(birds2,aes(y=ancestry,x=release_site,
                 group=interaction(as.factor(survival),release_site),
                 colour=release_site,shape=as.factor(survival)))+
  geom_point(size=3,alpha=0.8,
             position=position_jitterdodge(
               jitter.width =0.2,jitter.height=0,dodge.width=0.9))+
  scale_shape_manual(values=c(1,19),name="",labels=c("Not survived","Survived"))+
  scale_color_manual(values=natparks.pals("Banff")[c(4,7,1)],
                     guide="none")+
  geom_violin(fill=NA,aes(group=interaction(survival,release_site)),size=0.8)+
  ylab("Ancestry")+xlab("Release site")+ylim(0,1)+
  scale_x_discrete(labels=c("Lillooet","Pemberton","Squamish"))
  
gg1

#ggsave("C:/Users/Steph/TAMU_OneDrive/Thrushes/translocation/Fig2.png",
#       width=6,height=4)

```

* Next, points coloured by model-fitted survival probabilities

```{r,fig.width=5,fig.height=4}


ggplot(phi,aes(x=release_site,y=ancestry,colour=est.adj))+
  geom_jitter(size=4,shape=19,alpha=0.8,width=0.1)+
  viridis::scale_colour_viridis(option="mako",end=0.9,name="survival\nprobability")


```

```{r,fig.width=6,fig.height=4}


ggplot(phi_pemberton,aes(x=release_site,y=ancestry,colour=est.adj))+
  geom_jitter(size=4,shape=19,alpha=0.8,width=0.2)+
  viridis::scale_colour_viridis(option="mako",end=0.9,name="survival\nprobability")


```